How to improve your relationship with your future self

### 2016-11-29 responsetoreviewers.md

Dear Cristian and Diego,

Thanks for giving us the chance to revise the paper on workflow.

In response to your comments and those of the reviewer, we have added XX
mentions about experiments, although we realize that the issue is really about
transparency and not experiments in particular.

About the comment that much of the material appeared in an non-peer reviewed
journal previously: When I was approached to participate in this journal, wI
volunteered to send this previously non-peer reviewed article as a submission
to be peer-reviewed. That idea was accepted. As I worked to update the
previously published version, I invited Maarten Voors to work on the paper
with me. So the version that we sent differs a bit more from the 2011 version
than I had anticipated. I don't currently see us writing a whole new paper on
this topic. We really appreciate that you allowed our paper to go through the
peer-review process and very much hope that the version that we attach here is
satisfactory. It is still more of a version 2.0 than a completely new set of
thoughts on the topic. We are happy to specifically note the differences from
the first piece and have done so here.

We have attached the file here in pdf and LaTeX format. Don't hesitate to let
us know if you would like other revisions.

Best

Jake and Maarten

Regarding the comments of Reviewer 1:

We very much appreciate the thoughtful and supportive and helpful review. We
respond to the points made here.

 - We have now pointed out explicitly where we have updated and changed the
   paper to reflect our own changes thinking and the changes in the technology
landscape.

 - We have kept the primary lessons the same: we submitted this manuscript to
   RCP in order to gain more impact from the non-peer reviewed version, and
this idea was approved by the folks working on the project at that time. We
also feel like the primary lessons stand the test of time even as the specific
methods for achieving them might change.

 - The editors highlighted to us that the issue is not primarily about
   experiments but rather about transparency, so, although we have mentioned
experiments more in this version, we have not entirely changed the focus on
the piece.

 - We hope that we have clarified the role of Dropbox, and about makefiles and
   the make system.

 - We also have edited the paper to try to emphasize that there may be many
   best practices as long as they all are created to keep in mind the central
lessons of the paper. (For example, we believe that one can do transparent
scientific workflow using many different programming languages, computing
systems, and processes as long as the central lessons are followed.)

 - We have tried to quash the typos and formatting problems.

 - We have downplayed the parenting references.

Reviewer 1

This is a helpful paper that guides readers through a set of lessons to improve the way they organize their work.  If this piece appears in RCP, the authors should do more to highlight the new contributions over the very similar piece in The Political Methodologist. The primary lessons and many of the examples are identical or very similar. The paper should also be more closely tied to experimental work, rather than data analysis in general.  Are any of these problems specific to experiments?  Are there other issues that come up in experiments that are unique to that research environment?

The authors’ perspective on the role of Dropbox is unclear.  It seems that there are a lot of potential problems here.  In this way, the paper wavers from being about best practices to being about a diversity of approaches.  The authors may want to take a harder line, at the expense of trying to be comprehensive.  Also, if it mentions them, the paper should be more clear about the content and role of the “makefile” (p. 7) and “make” (p. 10).

The paper has many typos and mechanical problems like missing spaces, inconsistent formatting of “GitHub” and “LaTeX”, a missing figure number (“??”), inconsistent spacing around operators in code (see the tests on p.9 vs. the modeling on pp. 2 and 3), etc.  There are also some odd phrasings and interjections that I recommend removing (footnote 12, “Google those terms”, “People at UC Berkeley”, etc.  I appreciate and enjoyed the paper’s  casual tone, but it goes too far in my opinion. The parenting references, for example, are frequent and detailed enough to distract the reader.

### 2016-12-07 workflow.Rmd
made some edits

---
title: How to improve your relationship with your future self, V 2.0
author:
- name: Jake Bowers
  affiliation: <jwbowers@illinois.edu>,  University of Illinois at Urbana-Champaign, Departments of Political Science and Statistics; Fellow, White House Social and Behavioral Sciences Team.
- name: Maarten Voors
  affiliation: <maarten.voors@wur.nl>, Wageningen University, Development Economics Group.  \emph{Acknowledgements:} Many thanks to the EGAP Learning Days 2016 participants in Santiago Chile, to the BITSS team and the Department of Economics and Ted Miguel at UC Berkeley where Maarten was a visiting researcher during spring 2016. Maarten gratefully acknowledges financial support from N.W.O. grant 451-14-001. The previous version of this paper benefited from comments and discussions with Mark Fredrickson, Brian Gaines, Kieran Healy, Kevin Quinn, Cara Wong, Mika LaVaque-Manty and Ben Hansen. The source code for this document may be freely downloaded and modified from <https://github.com/jwbowers/workflow>.
bibliography: references.bib
csl: american-political-science-review.csl
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
    pdf_document:
      keep_tex: true
      number_sections: true
      fig_width: 7
      fig_height: 7
      fig_caption: true
      template: bowersarticletemplate.latex
      pandoc_args: [
      "-M", "secPrefix=\\S",
      "--filter", "/usr/local/bin/pandoc-crossref"
      ]
      md_extensions: +autolink_bare_uris+ascii_identifiers+tex_math_single_backslash
...

@@

> | Do I contradict myself?
| Very well then I contradict myself,
| (I am large, I contain multitudes.)
| [@whitman1855song]

@@

How long does it take from planning a study to publication? How much time passes to the first reproduction of a study? Is three years too short? Is ten years too long? We suspect that few of our colleagues in the social and behavioral sciences conceive of, field, write, and publish a data driven study faster than about three years. We also suspect that, if some other scholar decides to re-examine the analyses of a published study, it will occur after publication. Moreover, this new scholarly activity of learning from one anothers' data and analyses, can occur at anytime, many years past the initial publication of the article.^[The process of reproducing past findings can occur when one researcher wants to build on the work of another. It can also occur within the context of classes --- some professors assign reproduction tasks to students to aid learning about data analysis and statistics. In addition to those models, recently reproduction of research has been organized to enhance the quality of public policy in the field of economic development by the 3ie Replication Program (@brown2014quality) and to assess the quality of scientific research at the level of subdisciplines in social psychology (@open2015estimating) and within experimental economics (@camerer2016evaluating). Another of 29 research teams recently collaborated on a project focusing on applied statistics to see if the same answers would emerge from re-analyses of the same data set (@silberzahn2015crowdsourced).]

If we cannot count on our memories about why we made such and such a design or analysis decision, then what should we do? How can we minimize regret with our past decisions? How can we improve our relationship with our future self?  This essay is a heavily revised and updated version of @bowers2011tpm and provides some suggestions for practices that will make reproducible data analysis easy and quick. Specifically, this piece aims to amplify some of what we already ought to know and do, and highlight some current practices, platforms, and possibilities.^[@king1995replication and @nagler1995coding were two of the first pieces introducing these kinds of ideas to political scientists. Now, the efforts to encourage transparency and ease of learning from the data and analyses of others has become institutionalized with the DA-RT inititiave <http://www.dartstatement.org/> @lupia2014openness. These ideas are spreading beyond political science as well @freese2007replication and @asendorpf2013recommendations, <http://osf.io> and <http://www.bitss.org/>.] We aim to provide practical advice about how to *do work* such that one complies with such recommendations as a matter of course and, in so doing, can focus personal regret on bad past decisions that do not have to do with data analysis and the production of scholarly papers.

@@

All results (numbers, comparisons, tables, plots, figures) should arise from
code not from a series of mouse clicks or copying and pasting.^[To our future
selves: both of us are using a computer control device called a track-pad, and
haven't used an older device called a
[mouse](https://en.wikipedia.org/wiki/Computer_mouse) in some years. We no
longer hear a clicking sound when we operate the control device either. We are
not sure why we talked about mouse clicks just now.] If you wanted to re-create
the figure from say 2011 but include a new variable or specification, you
should be able to do so with just a few edits to the code rather than knowledge
of how you used a pointing device in your graphical user interface all those
years ago.

For example, using R [@R:2011], a open-source statistical software program, you
might specify that a file, called `fig1.pdf` was produced by the following set
of commands in a file called `makefig1.R`. Lets look at some annotated R code
you could produce:

@@

Now, in the future if you wonder how 'that plot on page 10' was created, you
will know: (1) 'that plot' is from a file called `fig1.pdf`and (2)
`fig1.pdf`was created in `makefig1.R`. Any time in a future, provided R still
exists, changing the figure will require some quick edits of commands already
written.^[We use data from @norris2015data throughout this paper.]  In a future where R does not exist, you (or someone else) will at
least be able to read the plain text R commands and use them to write code in a
new favorite statistical computing language: R scripts are written in [plain
text](http://en.wikipedia.org/wiki/Plain_text), and plain text is a format that
will be around as long as computer programmers write computer programs.^[Since
R is open source, you will also be able to download an old version of R,
download an old-fashioned open-source operating system (like Ubuntu 22), and
run the old-fashioned statistical computing environment in the old-fashioned
operating system in a virtual machine on your new-fashioned actual machine.]

Moreover, coding saves time. Often, a lot of time, especially if projects grow
in the number of files, partners, or complexity of analysis. Manually importing
one data file into R, may be effortless. Importing 100 files is another issue,
and the time costs of each manual action add up quickly (and the probability
for error increases).

Also, realize that *file names send messages* to your future self. This means
that if you name your files with evocative and descriptive names, your
collaborators are less likely to call you at midnight asking for help and you
will remove some regret from your future self and protect your friendships and
working relationships. For example, if you are studying inequality and protest,
you might try naming a file something like `inequality-and-protest-figures.R`
instead of `temp9.R` or `supercalifragilisticexpialidocious.R`.  By the way,
the extension `.R` tells us and the operating system that the file contains R
commands. This part of the filename enables us to quickly search our antique
hard drives for files containing R scripts.^[We think that some method of
tagging files by the purpose of the file will continue help analysts find and organize their files for long after the idea of a computer mouse ceases to make sense.]

Coding helps us avoid making mistakes. For example, in our example above we may
be interested in how many people protest. We may use a data file containing all
protests for several years. People often create a tabular display of this data
by copying the results manually to the working paper document. In copying we
can make mistakes. So a better approach is to ask your favorite data analysis
program to automatically create the table (in the format you like, with
horizontal lines, 3 decimals, a title, etc) and output this in whatever file
type you need (.pdf, .rtf, etc). Now when we obtain new data (for additional
years, or updated figures for existing data) a new table can be created
quickly, so we make less mistakes and save time!

@@

Data analysis involves a long series of decisions. Each decision requires a
justification.  Some decisions will be too small and technical for inclusion in
the published article itself. Still, these need to be documented in the code
itself [@nagler1995coding]. Paragraphs and citations in the publication will
justify the most important decisions but the code itself documents the smaller
but still consequential decisions. So, one must code to communicate with
yourself and others. There are two main ways to avoid forgetting the reasons
you did something with data: comment your code and tightly link your code with
your writing --- making your code literate.

@@

Group work requires version control. Many people are familiar with the 'track
changes' feature in modern WYSIWYG (what you see is what you get) word
processors or the fact that Dropbox allows one to recover previous versions of
files. These are both kinds of version control.  More generally, when we
collaborate, we'd like to do a variety of actions with our shared files.
Collaboration on data analytic projects is more productive and better when (1)
it is easy to see what has changed between versions of files; (2) members of
the team feel free to experiment and then to dump parts of the experimentation
in favor of previous work while merging the successful parts into the main body
of the paper; (3) the team can produce 'releases' of the same document
(one to MPSA, one to APSR, one to their parents) without spawning many possibly
conflicting copies of the same document; (4) people can work on the same files
at the same time without conflicting with one another, and can reconcile their
changes without too much confusion and clobbering. Clobbering is what happens
when your future self or your current collaborator saves an old version of a
file over a new version, erasing good work by accident.

Many rely on Dropbox, 'track changes' in Word, or Google Docs for
collaboration. Those tools tend to require you to communicate with other folks
in your group before you edit existing files. On Dropbox, only one of you can
edit and save a given file at a time. Tracking changes with Word tends to
function best if two people are not editing the same document (merging the
changes between two documents is not simple or builtin). Google Docs allows for
two people to edit the same document at the same time: but you must be online
with a fairly fast internet connection.^[Note to future selves: change this
section every five years or so. Stop saying 'internet' in 2026.] Communication
and trading off editing a file prevents your work (or your colleagues' work)
from getting lost when you both try to save the same file on top of each other. One may also use Dropbox as a kind of server for version
control (though the downside is if your collaborations or project grows large
(in terms of MBs) you need to switch to Dropbox's paid service). Maarten has been using
this for most of his collaborations. You can copy files from the Dropbox
directory into a local working directory so as to avoid clobbering and then
work on merging changes by hand before copying back to the Dropbox directory
and replacing existing files. You will however need to agree on an iron file
and directory naming rule with your collaborators to ensure that the files and directories are always named the same across machines and versions.

Another approach to organizing work on many files, including files for code and
for writing, is to use a formal [version
control](https://en.wikipedia.org/wiki/Version_control) strategy. As of 2016,
the standard for managing large collaborations with many files is the
[Git](https://en.wikipedia.org/wiki/Git_(software)) system. User friendly free
front-ends for Git currently make the process of learning and using Git very
convenient and integrated into the kind of workflow that your future self will
be proud of: [Open Science Framework](http://osf.io)
[bitbucket](http://bitbucket.com), [github](http://github.com). Jake uses
[github](http://github.com) for all of his collaborations with others as well
as with his future self: his collaborators appreciate some of the nice extra
features of github, such as the ability to keep a [shared task
list](https://guides.github.com/features/issues/). Learning about version
control systems takes a bit of time. We suggest, however, it is well worth the
time investment as it will save (lots!) of time later on.

@@

@@

If one has a moment to think in between teaching that new class, reading books
for an awards committee, evaluating application files for the admissions
committee, feeding popsicles to a sick child, and undertaking the odd bit of
research, one might say to oneself, "Before I write new code, I should write a
test of the code. I should write a little bit of code that lets me know that my
double-bootstrap procedure actually does what it is supposed to do."

@@

Lots of people are thinking about 'reproducible research', creating a
'reproducible workflow' and 'literate programming' these days. Google those
terms. Of course, the devil is in the details: Here we list a few of our own
attempts at enabling reproducible research. You'll find many other inspiring
examples on the web. Luckily, the open source ethos aligns nicely with academic
incentives, so we are beginning to find more and more people offering their
files online for copying and improvement. By the way, if you do copy and
improve, it is polite to alert the person from whom you made the copy about
your work and to credit them in some way.

@@

(1) We wrote this paper in the [R markdown](http://rmarkdown.rstudio.com/)
literate programming format using two different text editors (Jake used vim)
and (Maarten used [atom](https://atom.io/)). We organized our files and
collaboration on Github. The source code for this document can be accessed,
downloaded and modified from <https://github.com/jwbowers/workflow>. We also
used Github Issues to send notes to each other and maintain a task list. The
nice thing about github is that is enables you to make "releases" of a project
which enable you to smooth the reproduction of all of the products of your
research (simulations, data analyses, tables, figures, etc..).  Similar
features are offered by the [Open Science Framework (OSF)](https://osf.io/).
Maarten recently participated in a workshop on research transparency organized
by the [Berkeley Institute for Transparency in the Social
Sciences](www.bitss.org) in which all [course material](https://osf.io/qh2nr/)
was accessible through OSF.  OSF integrates with Github nicely and a key
benefit is that both systems have been created to remain for a long time
(unlike perhaps your university drive).  (2) For one paper Jake simply mixed R
and \LaTeX code in one document (called the "Sweave" format of literate
programming) and then added that document and data into into a compressed
archive [@bowers2005dataverse]; (3) For another more computing intensive paper
Jake assembled a set of files that enabled reproduction of results using the
`make` system [@bowers2008dataverse]; (4) Jake also tried the "compendium"
approach [@gentleman2005reproducible,@gentleman2007statistical] which embeds an
academic paper within the R package system [@bowers2011dataverse]; The benefit
of the compendium approach is that one is not required to have access to a
command line for `make`: the compendium is downloadable from within R using
`install.packages()` and is viewable using the `vignette()` function in any
operating system than runs R.^[Notice that Jake's reproduction archives and/or
instructions for using them are hosted on the
[Dataverse](http://thedata.org/book/learn-about-project), which is another
system designed to enhance academic collaboration across time and space.]  The
idea that one ought to be able to install and run and use an academic paper
just as one installs and uses statistical software packages is very attractive,
and we anticipate that it will become ever easier to turn papers into R
packages as creative and energetic folks turn their attention to the question
of reproducible research.^[See for example <http://r-pkgs.had.co.nz/>] Finally,
(5) Maarten has been using Dropbox for most of his collaborative projects. See
comments above. As a tool for making files publicly available Dropbox is less
useful. You can of course make replication files available through creating an
public folder, but you will still need to refer to this folder on a particular
project website (an example is [Maarten's project on conflict and football](
https://www.dropbox.com/sh/wqosq99019mf04d/AAASTgJLkn6lmZJI_1veUK7ca?dl=0)).

There are some noteworthy new developments also. People at UC Berkeley have
been developing a web app, Jupyter Notebook (http://jupyter.org/), that enables
you to make and share documents that include text, code, equations and
visualizations in one [literate programming
environment](http://blog.jupyter.org/2015/05/07/rendering-notebooks-on-github/).
For researchers working across different platforms, there is
[Docker](https://www.docker.com/) which promises to enable all of the
collaborators on one project to use the same computing environment even if they
are using different laptops running different operating systems.

There are also are great notes programs that help you communicate with collaborators. Remember, while email is fantastic for fast communication, it is not a tool designed for project management. It may work for small projects with few co-authors but when the number of collaborators increase it becomes very difficult to keep track. There are some great online systems for task management, most of which have mobile apps also, examples include [Flow](http://www.getflow.com), [Asana](http://www.asana.com), [Wrike](http://www.wrike.com), [Basecamp](http://basecamp.com) and many others. There also are great programs for notes. For example, Simplenote is very basic but allows you to share and edit notes. Others include [Evernote](www.evernote.com) and for Windows users there is OneNote.

@@

# Remember that research ought to be credible communication.

@@

We all always collaborate. Many of us collaborate with groups of people at one
moment in time as we race against a deadline. All of us collaborate with
ourselves over time.^[What is a reasonable time-span for which to plan
for self-collaboration on a single idea?  Ask your advisers how long it took
them to move from idea to dissertation to publication.]  The time-frames over
which collaboration are required --- whether among a group of people working
together or within a single scholar's productive life or probably both --- are
much longer than any given version of any given software will easily exist.
Plain text is the exception. Thus, even as we extol version control systems,
one must have a way to ensure future access to them in a form that will still
be around when sentient cockroaches finally join political science departments
(by then dominated by cetaceans after humans are mostly uploads).^[The arrival
of the six-legged social scientists revives Emacs and finally makes Ctrl-a
Ctrl-x Esc-x Ctrl-c a [reasonable key
combination](http://kieran.healy.usesthis.com/). So far, in terms of version
control systems, Jake has used RCS, CVS, bazaar, subversion, and now mostly
uses git.]

But what if no one ever hears of your work, or, by some cruel fate, your
article does not spawn debate? Why then would you spend time to communicate
with your future self and others? Our own answer to this question is that we
want our work to be credible and useful to ourselves and other scholars
regardless.  What we report in our data analyses should have two main
characteristics: (1) the findings of the work should not be a matter of
opinion; and (2) other people should be able to reproduce the findings. That
is, the work represents a shared experience --- and an experience shared
without respect to the identities of others (although requiring some common
technical training and research resources).

@@

In the end, following these practices and those recommended by
@fredrickson2011tpm and @healy2011tpm among others working on these topics
allows your computerized analyses of your collections of stuff to be credible.
If then someone quibbles with your analyses, your future self can shoot them
the archive required to reproduce your work.^[Since you used plain text, the
files will still be intelligible, analyzed using commented code so that folks
can translate to whatever system succeeds R, or since you used R, you can
include a copy of R and all of the R packages you used in your final analyses
in the archive itself. You can even throw in a copy of whatever version of
linux you used and an open source virtual machine running the whole environment
using say, docker.] You can say, 'Here is everything you need to reproduce my
work.' To be extra helpful you can add 'Read the README file for further
instructions.' And then you can get on with your life: maybe the next great
idea will occur when your 4-year-old asks a wacky question after stripping and
painting her overly cooperative 1-year-old brother purple, or teaching a class,
or in a coffee shop, or on a quiet walk.

>>>>

---
title: How to improve your relationship with your future self
author:
- name: Jake Bowers
  affiliation: <jwbowers@illinois.edu>,  University of Illinois at Urbana-Champaign, Departments of Political Science and Statistics; Fellow, White House Social and Behavioral Sciences Team.
- name: Maarten Voors
  affiliation: <maarten.voors@wur.nl>, Wageningen University, Development Economics Group.  \emph{Acknowledgements:} Many thanks to the EGAP Learning Days 2016 participants in Santiago Chile, to the BITSS team, the Department of Economics and Ted Miguel at UC Berkeley where Maarten was a visiting researcher during spring 2016. Maarten gratefully acknowledges financial support from N.W.O. grant 451-14-001. A previous version of this paper benefited from comments and discussions with Mark Fredrickson, Brian Gaines, Kieran Healy, Kevin Quinn, Cara Wong, Mika LaVaque-Manty and Ben Hansen. The source code for this document may be freely downloaded and modified from <https://github.com/jwbowers/workflow>. This paper extends a previous version by Jake @bowers2011tpm. While some text remains the same we have expanded and updated the essay to reflect current developments.
bibliography: references.bib
csl: american-political-science-review.csl
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
    pdf_document:
      keep_tex: true
      number_sections: true
      fig_width: 7
      fig_height: 7
      fig_caption: true
      template: bowersarticletemplate.latex
      pandoc_args: [
      "-M", "secPrefix=\\S",
      "--filter", "/usr/local/bin/pandoc-crossref"
      ]
      md_extensions: +autolink_bare_uris+ascii_identifiers+tex_math_single_backslash
...

@@

@@

How long does it take from planning a study to publication? How much time passes to the first reproduction of a study? Is three years too short? Is ten years too long? We suspect that few of our colleagues in the social and behavioral sciences conceive of, field, write, and publish a data driven study faster than about three years. We also suspect that, if some other scholar decides to re-examine the analyses of a published study, it will occur after publication. Moreover, this new scholarly activity of learning from one anothers' data and analyses, can occur at anytime, many years past the initial publication of the article.^[The process of reproducing past findings can occur when one researcher wants to build on the work of another. It can also occur within the context of classes --- some professors assign reproduction tasks to students to aid learning about data analysis and statistics. In addition to those models, recently reproduction of research has been organized to enhance the quality of public policy in the field of economic development by the 3IE Replication Program (@brown2014quality) and to assess the quality of scientific research at the level of sub-disciplines in social psychology (@open2015estimating) and within experimental economics (@camerer2016evaluating). In another study 29 research teams recently collaborated on a project focusing on applied statistics to see if the same answers would emerge from re-analyses of the same data set (@silberzahn2015crowdsourced).]

If we cannot count on our memories about why we made such and such a design or analysis decision, then what should we do? How can we minimize regret with our past decisions? How can we improve our relationship with our future self?  This essay is a heavily revised and updated version of @bowers2011tpm and provides some suggestions for practices that will make reproducible data analysis easy and quick. Specifically, this piece aims to amplify some of what we already ought to know and do, and highlight some current practices, platforms, and possibilities.^[@king1995replication and @nagler1995coding were two of the first pieces introducing these kinds of ideas to political scientists. Now, the efforts to encourage transparency and ease of learning from the data and analyses of others has become institutionalized with the DA-RT initiative <http://www.dartstatement.org/> @lupia2014openness. These ideas are spreading beyond political science as well @freese2007replication and @asendorpf2013recommendations, <http://osf.io> and <http://www.bitss.org/>.] We aim to provide practical advice about how to *do work* such that one complies with such recommendations as a matter of course and, in so doing, can focus personal regret on bad past decisions that do not have to do with data analysis and the production of scholarly papers.

@@

All your results (numbers, comparisons, tables, plots, figures) should be produced from
code, not from a series of mouse clicks or copying and pasting.^[To our future
selves: both of us are using a computer control device called a track-pad, and
haven't used an older device called a
[mouse](https://en.wikipedia.org/wiki/Computer_mouse) in some years. We no
longer hear a clicking sound when we operate the control device either. We are
not sure why we talked about mouse clicks just now.] Imagine you wanted to re-create
a figure and include a new variable, you should be able to do so with just a few edits to the code rather than knowledge
of how you used a pointing device in your graphical user interface all those
years ago.

Lets look at an example. Using R [@R:2011], a open-source statistical software program, you
might specify that a file, called `fig1.pdf` is produced by the following set
of commands in a file called `makefig1.R`. Lets look at some annotated R code:

@@

Now, in the future if you wonder how 'that plot on page 10' was created, you
will know: (1) 'that plot' is from a file called `fig1.pdf`and (2)
`fig1.pdf`was created in `makefig1.R`. Any changes to the file in the future (provided R still
exists), will just require some quick edits of commands already
written.^[We use data from @norris2015data throughout this paper.]  Even if in the future R ceases to exist, you (or someone else) will at
least be able to read the plain text commands and use them to write code in a
new favorite statistical computing language: R scripts are written in [plain
text](http://en.wikipedia.org/wiki/Plain_text), and plain text is a format that
will be around as long as computer programmers write computer programs.

Moreover, coding saves time. Often, a lot of time, especially if projects grow
in the number of files, partners, or complexity of analysis. Manually importing
one data file into R, may be effortless. Importing 100 files is another issue,
and the time costs of each manual action add up quickly (and the probability
for error increases too).

Also, realize that *file names send messages* to coauthors and to your future self. This means
that if you name your files with evocative and descriptive names, your
collaborators are less likely to call you at midnight asking for help and you
will remove some regret from your future self and protect your friendships and
working relationships. For example, if you are studying inequality and protest,
you might try naming a file something like `inequality-and-protest-figures.R`
instead of `temp9.R` or `supercalifragilisticexpialidocious.R`.  By the way,
the extension `.R` tells us and the operating system that the file contains R
commands. This part of the filename enables us to quickly search our antique
hard drives for files containing R scripts.^[We think that some method of
tagging files by the purpose of the file will continue help analysts find and organize their files for long after the idea of a computer mouse ceases to make sense.]

Coding helps us to avoid making mistakes. For example, in our example above we may
be interested in how many people protest. We may use a data file containing all
protests for several years. People often create a tabular display of this data
by copying the results manually to the working paper document. In copying we
can make mistakes. So a better approach is to automatically create the table (in the format you like, with
horizontal lines, 3 decimals, a title, etc) and output this in whatever file
type you need (.tex. .pdf, .rtf, etc). Now when we obtain new data a new table can be created
quickly, so we make less mistakes and save time!

@@

Data analysis involves a long series of decisions. Each decision requires a
justification.  Some decisions will be too small and technical for inclusion in
the published article itself. Still, these need to be documented in the code
itself [@nagler1995coding]. Paragraphs and citations in the publication will
justify the most important decisions but the code itself documents the smaller
but still consequential decisions. So, one must code to communicate with
yourself and others. There are two main ways to avoid forgetting the reasons
you did something with data: comment your code and tightly link your code with
your writing -- making your code literate.

@@

Group work requires version control. Many people are familiar with the 'track
changes' feature in modern WYSIWYG (what you see is what you get) word
processors or the fact that Dropbox allows one to recover previous versions of
files. These are both kinds of version control.  More generally, when we
collaborate, we'd like to do a variety of actions with our shared files.
Collaboration on data analytic projects is more productive and better when (1)
it is easy to see what has changed between versions of files; (2) members of
the team feel free to experiment and then to dump parts of the experimentation
in favor of previous work while merging the successful parts into the main body
of the paper; (3) the team can produce 'releases' of the same document
(one to MPSA, one to APSR, one to their parents) without spawning many possibly
conflicting copies of the same document; (4) people can work on the same files
at the same time without conflicting with one another, and can reconcile their
changes without too much confusion and clobbering. Clobbering is what happens
when your future self or your current collaborator saves an old version of a
file over a new version, erasing good work by accident (or creating a conflicted copy resulting in an arduous back and forth between versions).

Using Dropbox, 'track changes' in Word, or Google Docs is one way to manage your collaborative efforts. They are popular tools, but
have some downsides as they require you to communicate with other folks
in your group before you can edit existing files. On Dropbox, you cannot
edit and save a given file at the same time. Tracking changes with Word has similar limitations (merging the
changes between two documents is not simple or builtin). Google Docs allows for
two people to edit the same document at the same time: but you must be online
with a fairly fast internet connection.^[Note to future selves: change this
section every five years or so. Probably stop saying 'internet' in 2026.] Communication
and trading off editing a file prevents your work (or your colleagues' work)
from getting lost when you both try to save the same file on top of each other. One may also use Dropbox as a kind of server for version
control. See example above. This system is fine untill your collaborations or project grows large
(in terms of MBs) and you may run out of storage space on your computer or you may need to upgrade your Dropbox's storage space). You will however need to agree on an iron rule for file and directory naming with your collaborators to ensure that the files and directories are always named the same across machines and versions (see above).

@@

However, best practice we feel is to use formal version control approaches to organizing work [version
control](https://en.wikipedia.org/wiki/Version_control) strategy. As of 2016,
the standard for managing large collaborations with many files is the
[Git](https://en.wikipedia.org/wiki/Git_(software)) system. User friendly free
front-ends for Git currently make the process of learning and using Git very
convenient and integrated into the kind of workflow that your future self will
be proud of: [Open Science Framework](http://osf.io)
[bitbucket](http://bitbucket.com), [github](http://github.com). Jake uses
[github](http://github.com) for all of his collaborations with others as well
as with his future self: his collaborators appreciate some of the nice extra
features of GitHub, such as the ability to keep a [shared task
list](https://guides.github.com/features/issues/). Learning about version
control systems takes a bit of time. We suggest, however, it is well worth the
time investment as it will save (lots!) of time later on.

@@

@@

Over the past decades more and more attention is being paid to 'reproducible research', creating a
'reproducible workflow' and 'literate programming' these days. As always, the devil is in the details: Here we list a few of our own
attempts at enabling reproducible research. You'll find many other inspiring
examples on the web. Luckily, the open source ethos aligns nicely with academic
incentives, so we are beginning to find more and more people offering their
files online for copying and improvement. By the way, if you do copy and
improve, it is polite to alert the person from whom you made the copy about
your work and to credit them in some way.

@@

(1) We wrote this paper in the [R markdown](http://rmarkdown.rstudio.com/)
literate programming format using two different text editors (Jake used vim)
and (Maarten used [atom](https://atom.io/)). We organized our files and
collaboration on GitHub. The source code for this document can be accessed,
downloaded and modified from <https://github.com/jwbowers/workflow>. We also
used GitHub Issues to send notes to each other and maintain a task list. The
nice thing about GitHub is that is enables you to make "releases" of a project
which enable you to smooth the reproduction of all of the products of your
research (simulations, data analyses, tables, figures, etc..).  Similar
features are offered by the [Open Science Framework (OSF)](https://osf.io/).
Maarten recently participated in a workshop on research transparency organized
by the [Berkeley Institute for Transparency in the Social
Sciences](www.bitss.org) in which all [course material](https://osf.io/qh2nr/)
was accessible through OSF.  OSF integrates with Github nicely and a key
benefit is that both systems have been created to remain for a long time
(unlike perhaps your university drive).  (2) For one paper Jake simply mixed R
and \LaTeX code in one document (called the "Sweave" format of literate
programming) and then added that document and data into into a compressed
archive [@bowers2005dataverse]; (3) For another more computing intensive paper
Jake assembled a set of files that enabled reproduction of results using the
`make` system [@bowers2008dataverse]; (4) Jake also tried the "compendium"
approach [@gentleman2005reproducible,@gentleman2007statistical] which embeds an
academic paper within the R package system [@bowers2011dataverse]; The benefit
of the compendium approach is that one is not required to have access to a
command line for `make`: the compendium is downloadable from within R using
`install.packages()` and is viewable using the `vignette()` function in any
operating system than runs R.^[Notice that Jake's reproduction archives and/or
instructions for using them are hosted on the
[Dataverse](http://thedata.org/book/learn-about-project), which is another
system designed to enhance academic collaboration across time and space.]  The
idea that one ought to be able to install and run and use an academic paper
just as one installs and uses statistical software packages is very attractive,
and we anticipate that it will become ever easier to turn papers into R
packages as creative and energetic folks turn their attention to the question
of reproducible research.^[See for example <http://r-pkgs.had.co.nz/>] Finally,
(5) Maarten has been using Dropbox for most of his collaborative projects. See
comments above. As a tool for making files publicly available Dropbox is less
useful. You can of course make replication files available through creating an
public folder, but you will still need to refer to this folder on a particular
project website (an example is [Maarten's project on conflict and football](
https://www.dropbox.com/sh/wqosq99019mf04d/AAASTgJLkn6lmZJI_1veUK7ca?dl=0)).

There are some noteworthy new developments also. See for example
a new web app, Jupyter Notebook (http://jupyter.org/), that enables
you to make and share documents that include text, code, equations and
visualizations in one [literate programming
environment](http://blog.jupyter.org/2015/05/07/rendering-notebooks-on-github/).
For researchers working across different platforms, there is
[Docker](https://www.docker.com/) which promises to enable all of the
collaborators on one project to use the same computing environment even if they
are using different laptops running different operating systems.

There are also are great notes programs that help you communicate with collaborators. Remember, while email is fantastic for fast communication, it is not a tool designed for project management. It may work for projects with few co-authors but when the number of collaborators increase it becomes very difficult to keep track. Online systems for task management, if which most have mobile apps, include amongst otheres [Flow](http://www.getflow.com), [Asana](http://www.asana.com), [Wrike](http://www.wrike.com), [Basecamp](http://basecamp.com), Simplenote, [Evernote](www.evernote.com) and for Windows users there is OneNote.

@@

# Research ought to be credible communication.

@@

We all always collaborate. Many of us collaborate with groups of people at one
moment in time as we race against a deadline. All of us collaborate with
ourselves over time. The time-frames over
which collaboration are required --- whether among a group of people working
together or within a single scholar's productive life or probably both --- are
much longer than any given version of any given software will easily exist.
Plain text is the exception.

But what if no one ever hears of your work, or, by some cruel fate, your
article does not spawn debate? Why then would you spend time to communicate
with your future self and others? Our own answer to this question is that it is efficient and that out of principle we
want our work to be credible and useful to ourselves and other scholars.  What we report in our data analyses should have two main
characteristics: (1) the findings of the work should not be a matter of
opinion; and (2) other people should be able to reproduce the findings. That
is, the work represents a shared experience --- and an experience shared
without respect to the identities of others (although requiring some common
technical training and research resources).

@@

In the end, following these practices and those recommended by
@fredrickson2011tpm and @healy2011tpm among others working on these topics
allows your computerized analyses of your collections of stuff to be credible.
If then someone quibbles with your analyses, your future self can shoot them
the archive required to reproduce your work.^[Since you used plain text, the
files will still be intelligible, analyzed using commented code so that folks
can translate to whatever system succeeds R, or since you used R, you can
include a copy of R and all of the R packages you used in your final analyses
in the archive itself. You can even throw in a copy of whatever version of
Linux you used and an open source virtual machine running the whole environment
using say, Docker.] You can say, 'Here is everything you need to reproduce my
work.' To be extra helpful you can add 'Read the README file for further
instructions.' And then you can get on with your life: maybe the next great
idea will occur when your 4-year-old asks a wacky question after stripping and
painting her overly cooperative 1-year-old brother purple, or teaching a class,
or in a coffee shop, or on a quiet walk.
